---
title: "Polygenic_Lab"
author: "John Novembre and Jennifer Blanc"
date: "6/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
```

# Polygenic evolution

## Introduction

In this lab, you will have a set of functions that allow for simulating multiple traits with a polygenic basis and possible pleiotropic effects of alleles.  You have the opportunity to explore the effects of different assumptions about the function relating phenotype to fitness, genetic covariance among traits, population sizes, and number of loci. First, familiarize yourself with the functions and core simulation code below. 

In terms of the model, a few technical points are worth making: 
- We ignore LD due to linkage. Specifically, we model the expected change in allele frequency at each locus using the calculated mean marginal fitness of each allele. ($E[p_{t+1}] =p_t +   p_t(1-p_t)(w_1-w_2)/\bar{w}$)
- We will have diploid individuals but not simulate the effects of dioecy, because we will  simulate new offspring as the random union of gametes from a large gamete pool.  Specifically, we draw genotypes $X_{i,t} \sim \mathrm{Bin}(2,p_t)$  (i.e. our model is more appropriate for potentially selfing hermaphrodites, and less for monoecious random mating populations).  

### Background functions for use in our simulation
```{r}
#####  FUNCTIONS FOR USE IN CODE #### 

# Produce phenotypes
set_phenotypes <- function(geno_mat, N, n_traits, add_effect_sizes, dom_effect_sizes, sig2_env_tmp = sig2_env){

  phi_mat_tmp <- matrix(rep(0,N*n_traits),nrow = N, ncol = n_traits)
  for(i in 1:N){
#    if(sig2_env_tmp>0)
      phi_mat_tmp[i,] <- rnorm(n_traits,baseline_phi_val+geno_mat[i,]%*%add_effect_sizes + (geno_mat[i,]==1)%*%dom_effect_sizes, sig2_env_tmp)
#    else 
#      phi_mat_tmp[i,] <- baseline_phi_val+geno_mat[i,]%*%add_effect_sizes + (geno_mat[i,]==1)%*%dom_effect_sizes
  }
  return(phi_mat_tmp)
}

# Calculate fitnesses : All phenotypes selected towarsd an optimum
calc_fitness_twovar_gaussian <- function (x, x0=opt_phi_val,scale=opt_phi_scale){
  fit = exp( -sum(((x-x0)/scale)^2) ) 
  return(fit)
}

# Calculate fitnesses : Only first phenotype selected towarsd an optimum
calc_fitness_onevar_gaussian <- function (x, x0=opt_phi_val,scale=opt_phi_scale){
  fit = exp( -sum(((x[1]-x0[1])/scale[1])^2) ) 
  return(fit)
}

# Calculate fitnesses : Neutral case
calc_fitness_neutral <- function (x){
  fit = rep(1,length(x[1]))
  return(fit)
}

calc_fitness_wrapper <- function (x,model=1){
  if(model == 0) 
    return( calc_fitness_neutral(x) )
  if(model == 1)
    return ( calc_fitness_onevar_gaussian(x) ) 
  if(model == 2)
    return (calc_fitness_twovar_gaussian (x) )
}

# Calculate average phenotypic effect sizes
calc_avg_add_phi_effect_sizes <- function(phi_mat,geno_mat,n_traits, L){
  avg_add_phi_effect_sizes <- matrix(rep(0,L*n_traits), nrow=L, ncol= n_traits)
  for(i in 1:L){
    for(j in 1:n_traits){
      avg_add_phi_effect_sizes [i,j] <- lm(phi_mat[,j]~geno_mat[,i])$coefficients[2]
    }
  }
  return(avg_add_phi_effect_sizes)
}

# Calculate average fitness effect sizes
calc_avg_add_fit_effect_sizes <- function(fit_vec,geno_mat){
  avg_add_fit_effect_sizes <- rep(0,L)
  for(i in 1:L){
    avg_add_fit_effect_sizes [i] <- lm(fit_vec~geno_mat[,i])$coefficients[2]
  }
  return(avg_add_fit_effect_sizes)
}

# Calculate marginal fitness per allele 
calc_mar_fitness_per_allele_mat <- function(geno_mat,fit_vec,L){
  mar_fitness_per_allele_mat <- matrix(rep(0,L*2),nrow=L,ncol=2)
  for(i in 1:L){
    mar_fitness_per_allele_mat[i,1] <- ((geno_mat[,i]==1)%*%fit_vec + 2*(geno_mat[,i]==2)%*%fit_vec) / sum((geno_mat[,i]==1)+2*(geno_mat[,i]==2))
    mar_fitness_per_allele_mat[i,2] <-((geno_mat[,i]==1)%*%fit_vec + 2*(geno_mat[,i]==0)%*%fit_vec) / sum((geno_mat[,i]==1)+2*(geno_mat[,i]==0))
  }
  return(mar_fitness_per_allele_mat)
}

## Update allele frequencies due to selection and drift
update_geno_mat <- function(geno_mat,mar_fitness_per_allele,fit_vec, N, L){
  
  geno_mat_new <- matrix(rep(0,N*L),nrow = N, ncol=L)
  w_bar <- mean(fit_vec)
  for(i in 1:L){
    p_t = mean(geno_mat[,i])/2
    if(p_t == 1 || p_t == 0) {
      delta_p = 0
    } else
      delta_p = p_t * (1-p_t) * (mar_fitness_per_allele[i,1]-mar_fitness_per_allele[i,2]) / w_bar  
    p_det = p_t + delta_p
    if(p_det > 1) p_det = 1
    if(p_det < 0) p_det = 0
    geno_mat_new[,i] <- rbinom(n=N,size=2,prob=p_det)
  }
  return(geno_mat_new)
}
```

### Simulation driver code  

Set up default parameters.
```{r}
## Set key parameters that determine scale of the simulation
L <- 100  # Number of loci
time_steps <- 100  #100  # Maximum number of time steps
n_traits <- 2
N <- 200
fitness_mod <- 1  #  0 : neutral; 1 : selection on first trait; 2: selection on two traits
# Params of the trait model
n_traits <- 2 # Number of traits
per_locus_effect_size_scale <- 2/L   #(Max: genetic var = 2 * 0. 5 * x * L)
sig2_env <- rep(1,n_traits) * 0  # Environmental variance   # L 
record_Gmat <- 0  # record G matrix per iteration.  If set to 0 runs faster

# Baseline phenotype and optimum 
baseline_phi_val <- rep(0,n_traits)   # Baseline phenotype val that deviations are taken from
opt_phi_val <- rep(5,n_traits)   # Optimum phenotype value
opt_phi_scale <- rep(2.5^2,n_traits)  # Optimum scale (i.e. sig2 in Gaussian selection model)

# Effect sizes
effect_sizes_mean_vec <- rep(0, n_traits) # mean effects sizes
effect_size_var_vec <- rep(per_locus_effect_size_scale , n_traits)  # variance for effect sizes 
effect_size_covar <- -0.5 * per_locus_effect_size_scale # covariance in effect sizes (note: right now just one val for all pairs)
effect_sizes_var_mat <- matrix( rep(effect_size_covar,n_traits *n_traits), nrow=n_traits, ncol=n_traits)
diag(effect_sizes_var_mat) <- effect_size_var_vec 

# Frequencies drawn from a parameter
init_p_beta_param <- 1  # Parameter for symmetric beta distribution used to init allele frequencies
```

Initialize some key storage vectors. 
```{r}
## Set up vectors for storing statistics
time_max <- time_steps
p_time_mat <- matrix(rep(0,L*time_max),nrow =L, ncol = time_max)  # Init matrix to store allele freq
phi_bar_time_mat <- matrix(rep(0,n_traits*time_max),nrow = n_traits, ncol = time_max) # Init matrix to store  mean phi value
fit_bar_time_vec <- rep(0,time_max) # Init vec for mean fitness
beta1_time_vec <- rep(0,time_max) # Init vec for selection differential beta1
beta2_time_vec <- rep(0,time_max) # Init vec for selection differential beta2
beta12_time_vec <- rep(0,time_max) # Init vec for selection differential beta12
phi_var_time_mat <- matrix(rep(0,n_traits*time_max),nrow = n_traits, ncol = time_max) # Init matrix to store  mean phi value
phi_corr_time_vec <- rep(0, time_max) # Init vector for phenotypic covariance
fit_var_time_vec <- rep(0,time_max) # Init matrix for mean fitness
S_time_mat<- matrix(rep(0,n_traits*time_max),nrow = n_traits, ncol=time_max)
Gmat_time_mat <- matrix(rep(0,n_traits*n_traits*time_max),nrow= n_traits*n_traits,ncol=time_max)
```


## Simulation function  
```{r}
run_sim <- function(N,L,n_traits,per_locus_effect_size_covar,fitness_mod) {
  
  p_time_mat <- matrix(rep(0,L*time_max),nrow =L, ncol = time_max)  # Init matrix to store allele freq
  per_locus_effect_size_scale <- 2/L  # (Max: genetic var = 2 * 0. 5 * x * L)
  sig2_env <- rep(1,n_traits) * 0.1  # Environmental variance   # L 
  record_fiteffect <- 0 # Record fitness effects
  record_Gmat <- 0  # record G matrix per iteration.  If set to 0 runs faster

  # Baseline phenotype and optimum 
  baseline_phi_val <- rep(0,n_traits)   # Baseline phenotype val that deviations are taken from
  opt_phi_val <- rep(5,n_traits)   # Optimum phenotype value
  opt_phi_scale <- rep(2.5^2,n_traits)  # Optimum scale (i.e. sig2 in Gaussian selection model)

  # Effect sizes
  effect_sizes_mean_vec <- rep(0, n_traits) # mean effects sizes
  effect_size_var_vec <- rep(per_locus_effect_size_scale , n_traits)  # variance for effect sizes 
  effect_size_covar <- per_locus_effect_size_covar * per_locus_effect_size_scale # covariance in effect sizes (note: right now   just one val for all pairs)
  effect_sizes_var_mat <- matrix( rep(effect_size_covar,n_traits *n_traits), nrow=n_traits, ncol=n_traits)
  diag(effect_sizes_var_mat) <- effect_size_var_vec 

  # Frequencies drawn from a parameter
  init_p_beta_param <- 1  # Parameter for symmetric beta distribution used to init allele frequencies

  # NOTE: One can try here to take an alternative approach and draw effect sizes and allele 
  # frequencies that are correlated

  ##### Initialize effect sizes, allele frequencies, phenotype values, and genotype matrix ######

  p0 <- rbeta(L,init_p_beta_param,init_p_beta_param)  # Init allele freq's from a beta dist'n 
  add_effect_sizes <- mvrnorm(L,effect_sizes_mean_vec, effect_sizes_var_mat) # Additive effect sizes from multivariate norm
  dom_effect_sizes <- cbind(rep(0,L),rep(0,L)) # stub here for dominance effects sizes - all set to zero at moment
  
  geno_mat_0 <- matrix(rep(0,N*L),nrow=N,ncol=L) # Init genotypes 
  for(i in 1:L){
    geno_mat_0[,i] <- rbinom(n=N,size=2,prob=p0[i])
  }

  ###### Evolutionary model ###### 
  ## Code for main simulation loop
  if(1){
    geno_mat_t <- geno_mat_0 # Init temporary genotype matrix
    print("Starting main simulation loop...")
    for(t in 1:time_steps){

      # Draw phenotypes,  calculate fitness, calculate marginal fitness per allele
      phi_mat_t <- set_phenotypes(geno_mat_t, N, n_traits, add_effect_sizes, dom_effect_sizes) 
      if (t == 1){
        phi_mat_0 <- phi_mat_t
      }
      fit_vec_t <- apply(phi_mat_t,1,calc_fitness_wrapper,model=fitness_mod)
      mar_fitness_per_allele_t <- calc_mar_fitness_per_allele_mat(geno_mat_t,fit_vec_t, L)
  

      # Record key values
      p_time_mat[,t]<- apply(geno_mat_t,2,mean)/2
      phi_bar_time_mat[,t] <- apply(phi_mat_t,2,mean)
      phi_var_time_mat[,t] <- apply(phi_mat_t,2,var)
      phi_corr_time_vec[t] <- cor(phi_mat_t[,1],phi_mat_t[,2])
      fit_bar_time_vec[t] <- mean(fit_vec_t)
      fit_var_time_vec[t] <- var(fit_vec_t)
     
      #phi_g_mat_t <- set_phenotypes(geno_mat_t,sig2_env_tmp = 0)
      beta1_time_vec[t] <- lm(fit_vec_t ~ phi_mat_t[,1])$coefficients[2]
      beta2_time_vec[t] <- lm(fit_vec_t ~ phi_mat_t[,2])$coefficients[2]
      beta12_time_vec[t] <- lm(fit_vec_t ~ phi_mat_t[,1]*phi_mat_t[,2])$coefficients[2]
    
      for(i in 1:n_traits){
        S_time_mat[i,t] <- cov(phi_mat_t[,i],fit_vec_t)
      }


      # Record genetic var per trait and covar
      if(record_Gmat){
        fit_effect_size_vec <- calc_avg_add_fit_effect_sizes(fit_vec_t,geno_mat_t)

        phi_effect_size_mat <- calc_avg_add_phi_effect_sizes(phi_mat_t,geno_mat_t)
        for(i in 1:n_traits){
          Gmat_time_mat[i,t] <- sum(2 * p_time_mat[,t]*(1-p_time_mat[,t])* phi_effect_size_mat[,i]^2, na.rm = TRUE)
        }
        Gmat_time_mat[3,t] <- sum(2 * p_time_mat[,t]*(1-p_time_mat[,t])* fit_effect_size_vec^2, na.rm = TRUE)
        Gmat_time_mat[4,t] <- cov(phi_mat_t[,1],fit_vec_t)/var(phi_mat_t[,1])
      }

      # Update genotype matirx
      geno_mat_t <- update_geno_mat(geno_mat_t,mar_fitness_per_allele_t,fit_vec_t, N, L)
  
    }
  }
  print("Done.")
  return(list(p_time_mat,phi_bar_time_mat,add_effect_sizes,beta1_time_vec))
}

```

# Lab narrative

## Initial experimentation and visualization

By changing the `fitness_mod` parameter between values of 0, 1, and 2 we can run the model for the neutral case, for selection on one-trait, and for the selection on two-traits.  Inspect the following two plots for each run. For all three simulations we have set the covariance between effect sizes to be zero, meaning there is no pleiotropy.   

**Run Simulations** 
```{r}
neutral_sim <- run_sim(N=200,L=100,n_traits=2,per_locus_effect_size_covar=0,fitness_mod=0)
one_trait <- run_sim(N=200,L=100,n_traits=2,per_locus_effect_size_covar=0,fitness_mod=1)
two_trait <- run_sim(N=200,L=100,n_traits=2,per_locus_effect_size_covar=0,fitness_mod=2)
```

**Plot** 
```{r}
# Neutral Simulation
plot(1:time_steps,neutral_sim[[1]][1,],ylim=c(0,1),type="n",xlab="Generations", ylab="Allele frequencies")
for(i in 1:L)
  points(1:time_steps,neutral_sim[[1]][i,],col=i,type='l')
plot(neutral_sim[[2]][1,],neutral_sim[[2]][2,],type="l",xlab="Mean Phenotype 1",ylab="Mean Phenotype 2")
```

```{r}
# One trait  
plot(1:time_steps,one_trait[[1]][1,],ylim=c(0,1),type="n",xlab="Generations", ylab="Allele frequencies")
for(i in 1:L)
  points(1:time_steps,one_trait[[1]][i,],col=i,type='l')
plot(one_trait[[2]][1,],one_trait[[2]][2,],type="l",xlab="Mean Phenotype 1",ylab="Mean Phenotype 2", ylim = c(-5,5))
```

```{r}
# Two traits  
plot(1:time_steps,two_trait[[1]][1,],ylim=c(0,1),type="n",xlab="Generations", ylab="Allele frequencies")
for(i in 1:L)
  points(1:time_steps,two_trait[[1]][i,],col=i,type='l')
plot(two_trait[[2]][1,],two_trait[[2]][2,],type="l",xlab="Mean Phenotype 1",ylab="Mean Phenotype 2")
```

### Question 1 

Inspect the two plots made for each simulation. Explain in your own words why we see each Mean Phenotype 1 vs Mean Phenotype 2 pattern.  

### Question 2 

Re-run the 1 trait simulation for per_locus_effect_size_covar=0.5 and per_locus_effect_size_covar=-0.5. Plot the Mean Phenotype 1 vs Mean Phenotype 2 pattern for each simulation and explain the observed pattern.  

```{r}
## Your code goes here 

```


## Effective size with and without selection

### Question 3  

Using the relationship $H_t = H_0 e^{-t/(2N_e)}$ compute an effective population size from the mean heterozygosity for both a neutral simulation and a one trait selection model at t=100.  Hint: The mean heterozygosity at time $t$ can be computed as: `mean(2*(p_time_mat[,t]*(1-p_time_mat[,t])))`  

**Run Simulation**  
```{r}
neutral_sim <- run_sim(N=200,L=100,n_traits=2,per_locus_effect_size_covar=0,fitness_mod=0)
sel_sim <- run_sim(N=200,L=100,n_traits=2,per_locus_effect_size_covar=0,fitness_mod=1)
allele_freq_neutral <- neutral_sim[[1]] # time_steps x L
allele_freq_sel <- sel_sim[[1]] 
```

**Calculate Ne** 
```{r}
## Your code goes here 

```

### Question 4  

Does selection change the decay-of-heterozygosity effective size? If so, why?


## Allele Frequency change 

### Question 5  

Polygenic selection occurs through small allele frequency changes spread out across many loci. Write a function that takes the allele frequency matrix output by our simulation and calculates the difference in allele frequency between t=100 and t=1.  

**Run Simulation**  
```{r}
sim <- run_sim(N=200,L=100,n_traits=2,per_locus_effect_size_covar=0,fitness_mod=1)
allele_freq <- sim[[1]] # L x time_step
effect_size <- sim[[3]][,1]
```

**Calculate allele frequency change** 
```{r}
## Your code goes here 

```

### Question 6  

Plot the additive effect size of each site vs allele frequency change. Explain the resulting patterns.   

```{r}
## Your code goes here 
```


